---
title: "Working with Eyeris Databases: A Complete Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Eyeris Databases: A Complete Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(eyeris)
library(dplyr)
library(DBI)
```

# Introduction

The eyeris package provides powerful database functionality that serves as an alternative to CSV file storage for your eye-tracking data. This guide will walk you through the complete database workflow, from creation to data extraction, and explain why you might want to choose database storage over traditional CSV files.

## Why Use Databases Instead of CSV Files?

### The Case for Database Storage

**1. Scalability and Performance**
- **Massive datasets**: When processing hundreds or thousands of subjects, CSV files can become unwieldy. A single database file can efficiently store data from thousands of participants.
- **Faster queries**: Instead of reading multiple CSV files and filtering in R, databases can filter and aggregate data at the storage level, dramatically reducing memory usage and processing time.
- **Concurrent access**: Multiple researchers can query the same database simultaneously without file locking issues.

**2. Cloud Computing Advantages** 
- **Reduced I/O costs**: On cloud platforms (AWS, GCP, Azure), reading many small CSV files incurs significant I/O costs. A single database file reduces these costs dramatically.
- **Bandwidth efficiency**: Transferring one database file is much faster than syncing thousands of CSV files.
- **Storage optimization**: Databases compress data more efficiently than individual CSV files.

**3. Data Integrity and Management**
- **ACID compliance**: Databases ensure data integrity during concurrent writes and system failures.
- **Schema validation**: Built-in data type validation prevents corruption.
- **Metadata storage**: Databases automatically track creation timestamps, data types, and relationships.

**4. Analysis Efficiency**
- **SQL queries**: Leverage the power of SQL for complex filtering, aggregation, and joins.
- **Memory efficiency**: Load only the data you need instead of entire datasets.
- **Cross-platform compatibility**: DuckDB databases work across R, Python, and other analysis platforms.

### When to Use CSV vs Database

**Use CSV files when:**
- Small datasets (< 10 subjects)
- Simple analysis workflows
- Need human-readable data formats
- Working with external tools that require CSV input

**Use databases when:**
- Large datasets (> 20 subjects)
- Complex queries across multiple data types
- Cloud computing environments
- Collaborative research projects
- Long-term data storage and archival

# Getting Started: Creating Your First Database

## Basic Database Creation

The simplest way to create an eyeris database is by enabling database output during the bidsify process:

```{r basic-creation}
# Load your eye-tracking data
eyeris_data <- load_asc("path/to/your/data.asc")

# Process your data with epochs and confounds
processed_data <- eyeris_data %>%
  glassbox() %>%
  epoch(
    events = "TRIAL_START_{trial_type}_{trial_number}",
    limits = c(-0.5, 2.0),
    label = "trial_epochs"
  ) %>%
  summarize_confounds()

# Create database alongside CSV files
bidsify(
  processed_data,
  bids_dir = "~/my_eyetracking_study",
  participant_id = "001",
  session_num = "01",
  task_name = "memory_task",
  csv_enabled = TRUE,    # Still create CSV files
  db_enabled = TRUE,     # Also create database
  db_path = "study_database"  # Creates study_database.eyerisdb
)
```

## Database-Only Workflow (Cloud Optimized)

For cloud computing or large-scale processing, you can skip CSV creation entirely:

```{r cloud-workflow}
# Cloud-optimized: Database only, no CSV files
bidsify(
  processed_data,
  bids_dir = "~/my_eyetracking_study", 
  participant_id = "001",
  session_num = "01", 
  task_name = "memory_task",
  csv_enabled = FALSE,   # Skip CSV creation
  db_enabled = TRUE,     # Database only
  db_path = "study_database"
)
```

## Batch Processing Multiple Subjects

Here's how to efficiently process multiple subjects into a single database:

```{r batch-processing}
# Define your subjects
subjects <- c("001", "002", "003", "004", "005")
data_dir <- "~/raw_eyetracking_data"
bids_dir <- "~/processed_study_data"

# Process all subjects into one database
for (subject_id in subjects) {
  cat("Processing subject", subject_id, "\n")
  
  # Load and process subject data
  subject_data <- load_asc(file.path(data_dir, paste0("sub-", subject_id, ".asc"))) %>%
    glassbox() %>%
    epoch(
      events = "STIMULUS_{condition}_{trial}",
      limits = c(-1, 3),
      label = "stimulus_response"
    ) %>%
    summarize_confounds()
  
  # Add to database (automatically handles subject cleanup)
  bidsify(
    subject_data,
    bids_dir = bids_dir,
    participant_id = subject_id,
    session_num = "01",
    task_name = "attention_task", 
    csv_enabled = FALSE,  # Database only for efficiency
    db_enabled = TRUE,
    db_path = "attention_study_db"
  )
}
```

# Connecting to and Exploring Databases

## Basic Database Connection

```{r connection}
# Connect to your database
con <- eyeris_db_connect(
  bids_dir = "~/processed_study_data",
  db_path = "attention_study_db"  # Will look for attention_study_db.eyerisdb
)

# Always disconnect when done
# (or use on.exit to ensure cleanup)
on.exit(eyeris_db_disconnect(con))
```

## Database Overview and Exploration

```{r exploration}
# Get a comprehensive summary of your database
summary_info <- eyeris_db_summary("~/processed_study_data", "attention_study_db")

# View the summary
summary_info$subjects      # All subjects in database
summary_info$data_types    # Available data types
summary_info$sessions      # Session information
summary_info$tasks         # Task names
summary_info$total_tables  # Total number of tables

# List all available tables
all_tables <- eyeris_db_list_tables(con)
print(all_tables)

# Filter tables by data type
timeseries_tables <- eyeris_db_list_tables(con, data_type = "timeseries")
confounds_tables <- eyeris_db_list_tables(con, data_type = "run_confounds")

# Filter tables by subject
subject_001_tables <- eyeris_db_list_tables(con, subject = "001")
```

# Data Extraction: From Simple to Advanced

## Simple Data Extraction

The `eyeris_db_collect()` function provides easy one-liner access to your data:

```{r simple-extraction}
# Extract ALL data for ALL subjects (returns a named list)
all_data <- eyeris_db_collect("~/processed_study_data", "attention_study_db")

# View available data types
names(all_data)

# Access specific data types
timeseries_data <- all_data$timeseries
events_data <- all_data$events
confounds_data <- all_data$run_confounds
```

## Targeted Data Extraction

```{r targeted-extraction}
# Extract data for specific subjects only
subset_subjects <- eyeris_db_collect(
  bids_dir = "~/processed_study_data",
  db_path = "attention_study_db",
  subjects = c("001", "002", "003")
)

# Extract specific data types only
behavioral_data <- eyeris_db_collect(
  bids_dir = "~/processed_study_data", 
  db_path = "attention_study_db",
  data_types = c("events", "epochs", "confounds_summary")
)

# Extract data for specific sessions and tasks
session_01_data <- eyeris_db_collect(
  bids_dir = "~/processed_study_data",
  db_path = "attention_study_db", 
  sessions = "01",
  tasks = "attention_task"
)
```

## Working with Binocular Data

```{r binocular-extraction}
# Extract data from both eyes
binocular_data <- eyeris_db_collect(
  bids_dir = "~/processed_study_data",
  db_path = "attention_study_db"
)

# The function automatically combines left and right eye data
# Check if you have binocular data
unique(binocular_data$timeseries$eye_suffix)  # Should show "eyeL" and "eyeR"

# Extract data for specific eye only
left_eye_data <- eyeris_db_collect(
  bids_dir = "~/processed_study_data",
  db_path = "attention_study_db", 
  eye_suffixes = "eyeL"
)
```

## Epoch-Specific Extraction

```{r epoch-extraction}
# Extract specific epoch data
trial_epochs <- eyeris_db_collect(
  bids_dir = "~/processed_study_data",
  db_path = "attention_study_db",
  data_types = c("epochs", "confounds_events", "confounds_summary"),
  epoch_labels = "stimulus_response"  # Match your epoch label
)

# Multiple epoch types
multiple_epochs <- eyeris_db_collect(
  bids_dir = "~/processed_study_data", 
  db_path = "attention_study_db",
  data_types = "epochs",
  epoch_labels = c("stimulus_response", "baseline_period")
)
```

## Output Format Options

```{r output-formats}
# Default: Returns named list of dataframes (one per data type)
list_format <- eyeris_db_collect(
  "~/processed_study_data",
  return_list = TRUE  # Default
)

# Access individual data types
pupil_data <- list_format$timeseries
trial_data <- list_format$epochs

# Alternative: Single combined dataframe with data_type column
combined_format <- eyeris_db_collect(
  "~/processed_study_data", 
  return_list = FALSE
)

# Filter by data type in combined format
pupil_subset <- combined_format %>% 
  filter(data_type == "timeseries")
```

# Advanced Database Operations

## Direct SQL Queries

For maximum flexibility, you can write custom SQL queries:

```{r sql-queries}
# Connect to database
con <- eyeris_db_connect("~/processed_study_data", "attention_study_db")

# Custom SQL query
custom_query <- "
  SELECT subject_id, session_id, task_name, 
         AVG(pupil_clean) as mean_pupil,
         COUNT(*) as n_samples
  FROM timeseries_001_01_attention_task_run01_eyeL 
  WHERE pupil_clean IS NOT NULL
  GROUP BY subject_id, session_id, task_name
"

results <- DBI::dbGetQuery(con, custom_query)
print(results)

# Complex cross-table queries
complex_query <- "
  SELECT e.subject_id,
         e.matched_event,
         e.text_unique,
         AVG(t.pupil_clean) as mean_pupil_in_epoch,
         c.blink_rate_hz
  FROM epochs_001_01_attention_task_run01_stimulus_response_eyeL e
  JOIN timeseries_001_01_attention_task_run01_eyeL t 
    ON e.subject_id = t.subject_id 
    AND t.time_orig BETWEEN e.epoch_start AND e.epoch_end
  JOIN run_confounds_001_01_attention_task_run01_eyeL c
    ON e.subject_id = c.subject_id
  GROUP BY e.subject_id, e.matched_event, e.text_unique, c.blink_rate_hz
"

complex_results <- DBI::dbGetQuery(con, complex_query)
```

## Reading Individual Tables

```{r individual-tables}
# Read a specific table directly
specific_table <- eyeris_db_read(
  con = con,
  table_name = "timeseries_001_01_attention_task_run01_eyeL"
)

# Read with filters
filtered_data <- eyeris_db_read(
  con = con,
  data_type = "events",
  subject = "001",
  session = "01", 
  task = "attention_task"
)

# Read epoch data with specific epoch label
epoch_data <- eyeris_db_read(
  con = con,
  data_type = "epochs", 
  subject = "001",
  epoch_label = "stimulus_response"
)
```

# Real-World Analysis Examples

## Example 1: Pupil Response Analysis Across Subjects

```{r analysis-example-1}
# Extract all timeseries data
pupil_data <- eyeris_db_collect(
  "~/processed_study_data",
  data_types = "timeseries"
)$timeseries

# Analyze pupil responses by subject and condition
pupil_summary <- pupil_data %>%
  filter(!is.na(pupil_clean)) %>%
  group_by(subject_id, session_id) %>%
  summarise(
    mean_pupil = mean(pupil_clean),
    sd_pupil = sd(pupil_clean),
    samples_per_subject = n(),
    .groups = 'drop'
  )

print(pupil_summary)

# Compare to loading individual CSV files (much slower!)
# csv_files <- list.files("~/processed_study_data", 
#                        pattern = "*timeseries*.csv", 
#                        recursive = TRUE, full.names = TRUE)
# csv_data <- map_dfr(csv_files, read_csv)  # Slow for many files!
```

## Example 2: Event-Related Analysis

```{r analysis-example-2}
# Extract epochs and events together
epoch_data <- eyeris_db_collect(
  "~/processed_study_data", 
  data_types = c("epochs", "events")
)

# Analyze pupil response time-locked to events
event_analysis <- epoch_data$epochs %>%
  group_by(subject_id, matched_event, time_rel) %>%
  summarise(
    mean_pupil = mean(pupil_clean, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Create time windows
  mutate(
    time_window = case_when(
      time_rel < 0 ~ "baseline",
      time_rel < 1 ~ "early_response", 
      time_rel < 2 ~ "late_response",
      TRUE ~ "recovery"
    )
  )

# Statistical analysis by time window
window_stats <- event_analysis %>%
  group_by(subject_id, matched_event, time_window) %>%
  summarise(
    mean_pupil = mean(mean_pupil, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  pivot_wider(names_from = time_window, values_from = mean_pupil) %>%
  mutate(
    response_amplitude = early_response - baseline,
    sustained_response = late_response - baseline
  )
```

## Example 3: Quality Control and Confounds Analysis

```{r analysis-example-3}
# Extract confounds data for quality control
confounds_data <- eyeris_db_collect(
  "~/processed_study_data",
  data_types = c("run_confounds", "confounds_summary")
)

# Identify subjects with poor data quality
quality_control <- confounds_data$run_confounds %>%
  group_by(subject_id, session_id) %>%
  summarise(
    mean_blink_rate = mean(blink_rate_hz, na.rm = TRUE),
    mean_prop_invalid = mean(prop_invalid, na.rm = TRUE), 
    mean_gaze_variance = mean(gaze_x_var_px, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    high_blink_rate = mean_blink_rate > 0.5,  # Threshold
    high_invalid_data = mean_prop_invalid > 0.3,
    high_gaze_variance = mean_gaze_variance > 10000,
    exclude_subject = high_blink_rate | high_invalid_data | high_gaze_variance
  )

# View subjects recommended for exclusion
exclude_list <- quality_control %>%
  filter(exclude_subject) %>%
  select(subject_id, session_id, exclude_subject)

print(exclude_list)
```

# Performance Comparison: Database vs CSV

## Speed and Memory Benchmarks

```{r performance-comparison}
# Benchmark database approach
system.time({
  db_data <- eyeris_db_collect(
    "~/processed_study_data",
    subjects = c("001", "002", "003", "004", "005"),
    data_types = "timeseries"
  )
})

# Benchmark CSV approach (commented out - would be much slower!)
# system.time({
#   csv_files <- list.files("~/processed_study_data", 
#                          pattern = "*timeseries*.csv", 
#                          recursive = TRUE, full.names = TRUE)
#   csv_data <- map_dfr(csv_files[1:5], read_csv)  # Only first 5 subjects
# })

# Memory usage comparison
object.size(db_data)  # Database extraction
# object.size(csv_data)  # CSV loading

# File size comparison
db_file_size <- file.size("~/processed_study_data/derivatives/attention_study_db.eyerisdb")
csv_total_size <- sum(file.size(list.files("~/processed_study_data", 
                                          pattern = "*.csv", 
                                          recursive = TRUE, 
                                          full.names = TRUE)))

cat("Database file size:", round(db_file_size / 1024^2, 2), "MB\n")
cat("Total CSV file size:", round(csv_total_size / 1024^2, 2), "MB\n")
cat("Storage efficiency:", round(db_file_size / csv_total_size * 100, 1), "% of CSV size\n")
```

# Best Practices and Tips

## Database Management

```{r best-practices}
# 1. Always use descriptive database names
bidsify(data, db_path = "study_name_pilot_2024")  # Good
# bidsify(data, db_path = "my-project")            # Default, not descriptive

# 2. Use database-only mode for large studies
bidsify(data, csv_enabled = FALSE, db_enabled = TRUE)  # Efficient

# 3. Create separate databases for different experiments
bidsify(data, db_path = "experiment_1_attention")
bidsify(data, db_path = "experiment_2_memory") 

# 4. Always disconnect from databases
con <- eyeris_db_connect("~/data", "study_db")
# ... do your work ...
eyeris_db_disconnect(con)

# Or use on.exit for automatic cleanup
process_data <- function() {
  con <- eyeris_db_connect("~/data", "study_db")
  on.exit(eyeris_db_disconnect(con))
  
  # Your analysis code here
  results <- eyeris_db_collect("~/data", "study_db")
  return(results)
}
```

## Cloud Computing Optimization

```{r cloud-optimization}
# For AWS, GCP, or other cloud platforms:

# 1. Use database-only workflow to minimize I/O costs
process_cloud_data <- function(subject_list, input_bucket, output_bucket) {
  for (subject in subject_list) {
    # Download raw data
    download_from_cloud(subject, input_bucket)
    
    # Process and add to database (no CSV files)
    eyeris_data <- load_asc(local_file) %>%
      glassbox() %>% 
      epoch(...) %>%
      summarize_confounds()
    
    bidsify(
      eyeris_data,
      bids_dir = "local_processing",
      participant_id = subject,
      csv_enabled = FALSE,  # Skip CSV for cloud efficiency
      db_enabled = TRUE,
      db_path = "cloud_study_db"
    )
    
    # Clean up local files
    unlink(local_file)
  }
  
  # Upload final database
  upload_to_cloud("cloud_study_db.eyerisdb", output_bucket)
}

# 2. Use database for distributed analysis
analyze_cloud_data <- function() {
  # Download only the database file
  download_from_cloud("cloud_study_db.eyerisdb")
  
  # Extract only the data you need
  analysis_data <- eyeris_db_collect(
    "local_processing",
    data_types = c("epochs", "confounds_summary"),
    subjects = target_subjects
  )
  
  # Run analysis on extracted subset
  results <- run_statistical_analysis(analysis_data)
  
  return(results)
}
```

## Error Handling and Debugging

```{r error-handling}
# Safe database operations with error handling
safe_extract <- function(bids_dir, db_path, ...) {
  tryCatch({
    data <- eyeris_db_collect(bids_dir, db_path, ...)
    return(data)
  }, error = function(e) {
    cat("Error extracting data:", e$message, "\n")
    
    # Check if database exists
    db_file <- file.path(bids_dir, "derivatives", paste0(db_path, ".eyerisdb"))
    if (!file.exists(db_file)) {
      cat("Database file not found:", db_file, "\n")
      return(NULL)
    }
    
    # Try connecting to database
    con <- tryCatch({
      eyeris_db_connect(bids_dir, db_path)
    }, error = function(e2) {
      cat("Cannot connect to database:", e2$message, "\n")
      return(NULL)
    })
    
    if (!is.null(con)) {
      # List available tables for debugging
      tables <- eyeris_db_list_tables(con)
      cat("Available tables:\n")
      print(tables)
      eyeris_db_disconnect(con)
    }
    
    return(NULL)
  })
}

# Usage
data <- safe_extract("~/my_study", "study_database", 
                    subjects = c("001", "002"), 
                    data_types = "timeseries")
```

# Migration and Interoperability

## Converting Existing CSV Data to Database

```{r csv-to-database}
# If you have existing CSV files and want to migrate to database
migrate_csv_to_database <- function(bids_dir, db_path) {
  
  # Find all CSV files
  csv_files <- list.files(bids_dir, pattern = "\\.csv$", 
                         recursive = TRUE, full.names = TRUE)
  
  # Connect to database
  con <- eyeris_db_connect(bids_dir, db_path)
  on.exit(eyeris_db_disconnect(con))
  
  for (csv_file in csv_files) {
    cat("Processing:", basename(csv_file), "\n")
    
    # Parse filename to extract metadata
    # This depends on your CSV naming convention
    filename_parts <- parse_bids_filename(basename(csv_file))
    
    # Read CSV data
    csv_data <- read.csv(csv_file)
    
    # Write to database
    write_eyeris_data_to_db(
      data = csv_data,
      con = con,
      data_type = filename_parts$data_type,
      sub = filename_parts$subject,
      ses = filename_parts$session,
      task = filename_parts$task,
      # ... other parameters
    )
  }
  
  cat("Migration complete!\n")
}
```

## Exporting Database Data Back to CSV

```{r database-to-csv}
# Export specific data back to CSV format if needed
export_database_subset <- function(bids_dir, db_path, output_dir) {
  
  # Extract data from database
  data <- eyeris_db_collect(bids_dir, db_path, 
                             subjects = c("001", "002"),
                             data_types = c("timeseries", "events"))
  
  # Create output directory
  dir.create(output_dir, recursive = TRUE)
  
  # Export each data type
  for (data_type in names(data)) {
    filename <- file.path(output_dir, paste0(data_type, "_subset.csv"))
    write.csv(data[[data_type]], filename, row.names = FALSE)
    cat("Exported:", filename, "\n")
  }
}
```

# Conclusion

The eyeris database functionality provides a powerful, scalable alternative to CSV files for eye-tracking data storage and analysis. Key advantages include:

- **Performance**: Faster queries and reduced memory usage
- **Scalability**: Handle hundreds of subjects efficiently  
- **Cloud optimization**: Reduced I/O costs and bandwidth usage
- **Data integrity**: ACID compliance and automatic validation
- **Flexibility**: SQL queries and cross-platform compatibility

Whether you're running a small pilot study or a large-scale multi-site experiment, eyeris databases can streamline your workflow and improve analysis efficiency. Start with the simple `eyeris_db_collect()` function for most use cases, and leverage direct SQL queries when you need maximum flexibility.

For questions or issues with database functionality, please refer to the eyeris documentation or open an issue on the GitHub repository.

# Session Information

```{r session-info}
sessionInfo()
```
