---
title: "Learning Preprocessing Pitfalls with Simulated Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning Preprocessing Pitfalls with Simulated Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
img:not(.logo) {
  background-color: white;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.dpi = 300
)
```

```{r setup}
library(eyeris)
```

## 🎓 Why simulate pupil data?

Pupillometry preprocessing is a *pipeline*: each step assumes the steps before
it have already run. Run them out of order and the signal can be silently
corrupted --- no error, just wrong numbers. With **real** data you can never see
this clearly, because you do not know the ground truth underneath the artifacts.

`eyeris` ships a small, self-contained **simulator** that solves this for
teaching and testing:

- `simulate_eyeris()` generates a synthetic pupil time series whose statistical
  and morphological characteristics resemble real EyeLink recordings, and wraps
  it in a fully valid `eyeris` object;
- `sim_params()` lets you dial every component of the signal up or down;
- `sim_step_showcase()` and `sim_pitfall()` render **before/after** comparisons
  that make the consequences of each step --- and of getting the *order* wrong
  --- impossible to miss.

Because the generator is seeded, every figure below is exactly reproducible, and
because we build the signal ourselves we always know the ground truth.

## 🧪 Generating a synthetic recording

`simulate_eyeris()` returns an ordinary `eyeris` object --- the same class
[`load_asc()`][eyeris::load_asc] produces --- so it flows through the entire
pipeline unchanged (`glassbox()`, `plot()`, `epoch()`, and friends).

```{r generate}
sim <- simulate_eyeris(seed = 1, params = sim_params(duration_secs = 30))
class(sim)
```

The synthetic pupil signal is composed additively from independently
toggle-able parts (see `?sim_params` for the full list and the exact math):

- a **tonic baseline** with optional slow **drift** and a slow **hippus**
  oscillation,
- canonical task-evoked **phasic** dilations (the Hoeks & Levelt Erlang response),
- **blinks** --- runs of missing data flanked by the rapid partial-occlusion
  spikes that surround real blinks,
- isolated **transient** spikes (tracker glitches),
- broadband measurement **noise**, and optional high-frequency **line noise**.

Here is a three-second window showing a blink (the gap, with its tell-tale
down-then-up flank spikes) sitting on the noisy tonic signal:

```{r raw-signal, fig.height = 4}
b <- sim$timeseries$block_1
sel <- b$time_secs >= 6 & b$time_secs <= 9
plot(b$time_secs[sel], b$pupil_raw[sel],
  type = "l", col = "grey30",
  xlab = "time (s)", ylab = "pupil (a.u.)",
  main = "A window of raw simulated pupil data"
)
```

Because it is a real `eyeris` object, the full pipeline just works:

```{r pipeline, eval = FALSE}
sim |>
  eyeris::deblink() |>
  eyeris::detransient() |>
  eyeris::interpolate() |>
  eyeris::lpfilt() |>
  eyeris::zscore()
```

## 🔍 What a single step does

`sim_step_showcase()` runs the canonical pipeline up to a step of your choice
and plots the before/after of each step on a window bracketing the artifact that
step targets. For example, deblinking removes the blink and its flank spikes:

```{r step-showcase, fig.height = 6}
invisible(sim_step_showcase("deblink", seed = 1, verbose = FALSE))
```

## 🔀 Ordering pitfalls

The real payoff is `sim_pitfall()`. It generates data tuned to expose a specific
ordering mistake, then runs the pipeline **the wrong way and the right way on the
exact same signal**, so the only difference is the order of the steps. It returns
a quantitative metric and (by default) plots a three-panel comparison: the
**raw input** on top (so you can see the artifact you started with), the
**wrong-order** result (tinted red, with the correct result overlaid as a dashed
reference), and the **correct-order** result (tinted green).

Each metric table reports the `wrong` and `right` values and whether a `lower`
or `higher` value is the desired outcome.

### 1. Interpolating before deblinking

Interpolation can only bridge the *missing core* of a blink; it faithfully
preserves the steep occlusion **flank spikes**. If you interpolate first, those
spikes survive as large spurious deflections that deblinking can no longer
remove.

```{r pitfall-interp, fig.height = 8.5}
res <- sim_pitfall("interp_before_deblink", seed = 1, verbose = FALSE)
res$metrics
```

The wrong order leaves a residual deflection roughly the size of the occlusion
spike; the right order leaves essentially nothing.

### 2. Filtering before interpolating (a guardrail)

`lpfilt()` requires a gap-free signal and **errors** on any missing sample. This
is a deliberate guardrail: interpolation must come first. `sim_pitfall()`
captures the error so you can inspect it.

```{r pitfall-lpfilt-interp, fig.height = 8.5}
res <- sim_pitfall("lpfilt_before_interp", seed = 1, verbose = FALSE)
conditionMessage(res$wrong)
```

### 3. Filtering before removing transients

`detransient()` identifies spikes by their high sample-to-sample **speed**. Once
the signal has been low-pass filtered, that tell-tale speed is gone, so the
detector can no longer flag the artifacts.

```{r pitfall-lpfilt-detransient, fig.height = 8.5}
res <- sim_pitfall("lpfilt_before_detransient", seed = 1, verbose = FALSE)
res$metrics
```

Run on the raw signal, `detransient()` flags every transient; run after
filtering, it flags essentially none.

### 4. Decimating without an anti-alias filter

Downsampling by simply keeping one sample out of every *k* folds
high-frequency content back into the pupil band as a spurious low-frequency
oscillation. `downsample()` applies an anti-alias filter first and avoids it;
the naive alternative does not.

```{r pitfall-alias, fig.height = 8.5}
res <- sim_pitfall("naive_downsample", seed = 1, verbose = FALSE)
res$metrics
```

The naive decimation leaves a large aliased peak in the band; the anti-aliased
`downsample()` removes it.

### 5. Standardizing without detrending

When a slow linear **drift** is present, z-scoring alone leaves the trend
embedded in the standardized signal. `detrend()` removes it first.

```{r pitfall-detrend, fig.height = 8.5}
res <- sim_pitfall("omit_detrend", seed = 1, verbose = FALSE)
res$metrics
```

The residual slope of the final signal is clearly nonzero without detrending and
approximately zero with it.

## 🎨 Customizing for your own lessons

Every pitfall accepts your own `sim_params()`, so you can exaggerate or soften an
effect for a specific audience. For example, make the drift steeper for a more
dramatic detrend demo:

```{r custom-params}
steep <- sim_params(duration_secs = 40, drift = TRUE, drift_slope = 20)
showcase_metric("omit_detrend", seed = 1, params = steep)
```

Or compute a metric without plotting at all --- handy for exercises and
automated problem sets:

```{r metric-only}
showcase_metric("interp_before_deblink", seed = 1)
```

## 📋 Recap

| Pitfall | Lesson |
|---|---|
| `interp_before_deblink` | Deblink before interpolating, or blink flank spikes survive. |
| `lpfilt_before_interp` | Interpolate before filtering (filtering errors on missing data). |
| `lpfilt_before_detransient` | Remove transients before filtering, or the detector is defeated. |
| `naive_downsample` | Decimate only *with* an anti-alias filter (use `downsample()`). |
| `omit_detrend` | Detrend a drifting signal before z-scoring. |

The canonical `eyeris` order --- the one `glassbox()` enforces for you --- is
`deblink → detransient → interpolate → lpfilt → (downsample) → (detrend) →
zscore`. When in doubt, let `glassbox()` handle the ordering; these simulations
are here to show you *why* that order exists.

<div class="alert alert-light">
💡 <strong>Note on realism.</strong> The exaggerated single-sample flank spikes
and the optional line-noise tone are synthetic teaching devices chosen to make
each lesson unmistakable, not claims about the precise physiology of EyeLink
pupil diameter recordings.
</div>
